---
title: "Retrieving Human Variant Alleles from Ensembl Using API"
author: "Mira Sohn" 
output: html_document
---



### Aims

#### - This workflow is aimed at retrieving human variant alleles from Ensembl by using the [Ensembl REST API (GET overlap/id/:id)](https://rest.ensembl.org/documentation/info/overlap_id)

#### - Reference: [The Ensembl Variant Effect Predictor (McLaren et al., 2016)](https://pubmed.ncbi.nlm.nih.gov/27268795)


### Conda environment

#### - This analysis was performed under [conda](https://conda.io/projects/conda/en/latest/index.html) environment (see the [recipe](https://github.com/Mira0507/TRF_demo/blob/master/conda_r.yml))


```{r global_options, include=FALSE}


knitr::opts_chunk$set(
    warning=FALSE,
    message=FALSE
    )


```

### Loading packages

```{r loading_packages}

library(tidyverse)
library(httr) 
library(jsonlite) 

```

### Importing gene ids 

#### - This demonstration uses ensembl human gene id 


```{r importing_geneid}

# Importing my genes to a vector 
geneid <- scan("geneid.txt", character(), quote="")

# Exploring the imported object 
geneid[1:10]
class(geneid)
length(geneid)



```


### Retrieving data

```{r count_variants}


# Assign the server for API 
server <- "https://rest.ensembl.org"


variants.df <- data.frame()

for (i in 1:length(geneid)) {


    # Retrieve data using ensembl gene ids 
    ext <- paste0("/overlap/id/",
                  geneid[i],
                  "?feature=variation")

    r <- GET(paste(server, ext, sep = ""), 
             content_type("application/json"))
     
    stop_for_status(r)

    # Extract the variant ids 
    df <- data.frame(t(sapply(content(r),c))) %>% 
        gather("Variable", "Variant") %>% 
        dplyr::filter(str_detect(Variant, "rs")) %>%
        mutate(GENEID=geneid[i]) %>%
        dplyr::select(-Variable) 

    # Combine the retrieved variants to variants.df by rbinding
    variants.df <- rbind(variants.df, df)

}

# Ensure that any duplicated rows are removed
variants.df <- variants.df[!duplicated(variants.df),] %>%
    mutate(Variant=unlist(Variant))

# Explore the output data frame
head(variants.df)
dim(variants.df)
glimpse(variants.df)

# Count the number of variant alleles per gene id 
variants.sum <- variants.df %>%
    group_by(GENEID) %>%
    summarize(Counts=n())

# Explore the output data frame
head(variants.sum)
dim(variants.sum)
glimpse(variants.sum)

# Save the output as csv files
write.csv(variants.df, "total_variants.csv")
write.csv(variants.sum, "counted_variants.csv")
```




### Session Info

```{r session_info}

sessionInfo()

```
